---
title: "Unified Quranic Syntax Adapted Database - UQSAD Analysis"
output: html_notebook
---

# Unified Quranic Syntax Adapted Database - UQSAD Analysis

This notebook analyzes multiple Quran databases from different Mushaf layouts and publishers.

## File Summary & Code Explanation

### **Purpose**

This R Notebook systematically connects to, explores, and compares 11 different Quran databases (SQLite format) from various Mushaf layouts and publishers. It provides comprehensive analysis including database structure exploration, sample data extraction, and comparative statistics.

### **Target Databases (11 Total)**

-   **01-11**: Various Mushaf layouts including Indo-Pak (13-16 lines), QPC versions (15 lines), Uthmani script, Digital Khatt, Nastaleeq, and Tajweed versions

### **Dependencies & Libraries**

-   **DBI**: Database interface for R - enables database connections and SQL operations
-   **RSQLite**: SQLite driver for R - specifically handles SQLite database connections
-   **dplyr**: Data manipulation grammar - provides functions like filter(), select(), mutate(), bind_rows()
-   **purrr**: Functional programming tools - offers map(), walk() and other iteration functions
-   **stringr**: String manipulation - handles text processing and pattern matching
-   **knitr**: Document generation - enables table formatting and report creation
-   **tibble**: Modern data frames - provides enhanced data frame creation and printing

### **Main Functionality**

1.  **Database Setup**: Defines paths and metadata for all 11 Quran Mushaf databases
2.  **Connection Management**: Creates safe connection functions with error handling
3.  **Structure Exploration**: Discovers tables and schemas in each database using SQLite PRAGMA commands
4.  **Data Sampling**: Extracts first 5 rows from every table for content preview
5.  **Comparative Analysis**: Counts and compares row numbers across databases
6.  **Resource Cleanup**: Properly closes all database connections

### **Key Features**

-   **Error Handling**: Uses tryCatch() to prevent crashes from corrupted databases
-   **Systematic Organization**: Numbers databases (01-11) for clear identification
-   **Comprehensive Coverage**: Analyzes every table in every database
-   **Resource Management**: Properly manages database connections and memory

## Setup and Libraries

```{r setup, message=FALSE, warning=FALSE}
# Load DBI package - provides database interface for R
library(DBI)
# Load RSQLite package - enables SQLite database connections in R
library(RSQLite)
# Load dplyr package - provides data manipulation functions (filter, select, mutate, etc.)
library(dplyr)
# Load purrr package - provides functional programming tools (map, walk, etc.)
library(purrr)
# Load stringr package - provides string manipulation functions
library(stringr)
# Load knitr package - enables table formatting and document generation
library(knitr)
# Load tibble package - provides modern data frame creation and printing
library(tibble)
```

## Database Paths

```{r database_paths}
base_path <- "../data/QUL-Hafs-Mushafs-Mapping"

# Create a structured tibble (data frame) containing information about all 11 databases
db_info <- tibble(
  # Create sequence of numbers from 1 to 11 for database identification
  id = 1:11,
  # Define the internal names for each database (without folder numbers)
  name = c(
    "indopak-13-lines-layout-qudratullah",  # Database 1: Indo-Pak script, 13 lines, Qudratullah layout
    "qpc-v1-15-lines",                      # Database 2: Quran Printing Complex version 1, 15 lines
    "uthmani-15-lines",                     # Database 3: Uthmani script, 15 lines
    "digital-khatt-15-lines",               # Database 4: Digital Khatt font, 15 lines
    "indopak-13-lines-taj-company",         # Database 5: Indo-Pak script, 13 lines, Taj Company
    "qpc-hafs-15-lines",                    # Database 6: QPC Hafs version, 15 lines
    "qpc-nastaleeq-15-lines",               # Database 7: QPC Nastaleeq script, 15 lines
    "qpc-v2-15-lines",                      # Database 8: Quran Printing Complex version 2, 15 lines
    "qudratullah-indopak-15-lines",         # Database 9: Qudratullah Indo-Pak, 15 lines
    "taj-indopak-16-lines",                 # Database 10: Taj Indo-Pak script, 16 lines
    "qpc-v4-tajweed-15-lines"               # Database 11: QPC version 4 with Tajweed, 15 lines
  ),
  # Generate folder names with zero-padded numbers (01-, 02-, etc.) matching file system
  folder = sprintf("%02d-%s.db", id, name),
  # Generate the actual database file names (adding .db extension)
  db_file = paste0(name, ".db"),
  # Create full file paths by combining base path, folder, and database file name
  full_path = file.path(base_path, folder, db_file),
  # Create display names with numbers for clear identification in output
  display_name = sprintf("%02d - %s", id, name)
)

# Print the complete database information table for verification
print(db_info)
```

## Database Connection Function

```{r connection_function}
# Define function to safely connect to a SQLite database
connect_to_db <- function(db_path) {
  # Check if the database file actually exists at the specified path
  if (file.exists(db_path)) {
    # If file exists, create and return a database connection using RSQLite
    dbConnect(RSQLite::SQLite(), db_path)
  } else {
    # If file doesn't exist, show warning message with the missing path
    warning(paste("Database not found:", db_path))
    # Return NULL to indicate connection failed
    NULL
  }
}

# Define function to get list of all tables in a database
list_tables <- function(conn) {
  # Check if connection object is valid (not NULL)
  if (!is.null(conn)) {
    # If valid connection, get and return list of table names
    dbListTables(conn)
  } else {
    # If invalid connection, return NULL
    NULL
  }
}

# Define function to get the schema (structure) information for a specific table
get_table_schema <- function(conn, table_name) {
  # Check if connection is valid AND the table exists in the database
  if (!is.null(conn) && table_name %in% dbListTables(conn)) {
    # Execute SQLite PRAGMA command to get table structure information
    # PRAGMA table_info() returns: column ID, name, type, not null, default value, primary key
    dbGetQuery(conn, paste("PRAGMA table_info(", table_name, ")"))
  } else {
    # If connection invalid or table doesn't exist, return NULL
    NULL
  }
}
```

## Explore Database Structures

```{r explore_databases}
# Initialize empty list to store database connections for later use
db_connections <- list()
# Initialize empty list to store schema information for all databases
db_schemas <- list()

# Loop through each database in our db_info table (from row 1 to total number of rows)
for (i in 1:nrow(db_info)) {
  # Extract the internal database name from current row
  db_name <- db_info$name[i]
  # Extract the full file path from current row
  db_path <- db_info$full_path[i]
  # Extract the display name (with numbers) from current row
  display_name <- db_info$display_name[i]
  
  # Print header showing which database we're currently processing
  cat("\n=== Database:", display_name, "===\n")
  # Print the file path for verification
  cat("Path:", db_path, "\n")
  
  # Check if the database file actually exists at the specified path
  if (file.exists(db_path)) {
    # If file exists, attempt to connect to the database
    conn <- connect_to_db(db_path)
    # Store the connection in our connections list using the database name as key
    db_connections[[db_name]] <- conn
    
    # Check if connection was successful (not NULL)
    if (!is.null(conn)) {
      # Get list of all tables in this database
      tables <- list_tables(conn)
      # Print number of tables found and their names
      cat("Tables found:", length(tables), "\n")
      print(tables)
      
      # Initialize nested list to store schema info for this database
      db_schemas[[db_name]] <- list()
      # Loop through each table in this database
      for (table in tables) {
        # Get detailed schema information for current table
        schema <- get_table_schema(conn, table)
        # Store schema info in our nested structure
        db_schemas[[db_name]][[table]] <- schema
        # Print header for table schema
        cat("\nSchema for table '", table, "':\n")
        # Print the complete schema information
        print(schema)
      }
    }
  } else {
    # If database file doesn't exist, print error message
    cat("Database file not found!\n")
  }
}
```

## Sample Data from Each Database

```{r sample_data}
# Loop through each database connection we established earlier
for (db_name in names(db_connections)) {
  # Get the connection object for current database
  conn <- db_connections[[db_name]]
  # Find the display name for current database by matching internal name
  display_name <- db_info$display_name[db_info$name == db_name]
  # Check if connection is valid (not NULL)
  if (!is.null(conn)) {
    # Print header showing which database we're sampling data from
    cat("\n=== Sample data from", display_name, "===\n")
    
    # Get list of all tables in current database
    tables <- dbListTables(conn)
    # Loop through each table in current database
    for (table in tables) {
      # Print header showing which table we're sampling
      cat("\nSample from table '", table, "':\n")
      # Use tryCatch to handle any errors that might occur during data retrieval
      tryCatch({
        # Execute SQL query to get first 5 rows from current table
        sample_data <- dbGetQuery(conn, paste("SELECT * FROM", table, "LIMIT 5"))
        # Print the sample data
        print(sample_data)
      }, error = function(e) {
        # If any error occurs, print error message instead of crashing
        cat("Error reading table:", e$message, "\n")
      })
    }
  }
}
```

## Database Comparison Analysis

```{r database_comparison}
# Initialize empty list to store comparison results for all databases
comparison_results <- list()

# Loop through each database connection we have established
for (db_name in names(db_connections)) {
  # Get connection object for current database
  conn <- db_connections[[db_name]]
  # Check if connection is valid (not NULL)
  if (!is.null(conn)) {
    # Get list of all tables in current database
    tables <- dbListTables(conn)
    
    # Loop through each table in current database
    for (table in tables) {
      # Use tryCatch to handle any errors during row counting
      tryCatch({
        # Create SQL query to count total rows in current table
        count_query <- paste("SELECT COUNT(*) as row_count FROM", table)
        # Execute the count query and extract the numeric result
        row_count <- dbGetQuery(conn, count_query)$row_count
        
        # Check if comparison_results variable exists (safety check)
        if (!exists("comparison_results")) {
          comparison_results <- list()
        }
        
        # If this is first time seeing this table name, initialize empty tibble
        if (is.null(comparison_results[[table]])) {
          comparison_results[[table]] <- tibble(database = character(), rows = numeric())
        }
        
        # Get display name for current database
        display_name <- db_info$display_name[db_info$name == db_name]
        # Add current database's row count to the comparison table
        comparison_results[[table]] <- bind_rows(
          comparison_results[[table]], 
          tibble(database = display_name, rows = row_count)
        )
      }, error = function(e) {
        # If error occurs during counting, print error message
        cat("Error getting row count for", table, "in", db_name, ":", e$message, "\n")
      })
    }
  }
}

# Print header for comparison summary
cat("=== Database Comparison Summary ===\n")
# Loop through each table name in our comparison results
for (table_name in names(comparison_results)) {
  # Print table name header
  cat("\nTable:", table_name, "\n")
  # Print the comparison table showing row counts across all databases
  print(comparison_results[[table_name]])
}
```

## Cleanup Connections

```{r cleanup}
# Loop through each database connection in our connections list
for (conn in db_connections) {
  # Check if connection is valid (not NULL) before attempting to close
  if (!is.null(conn)) {
    # Properly close the database connection to free up system resources
    dbDisconnect(conn)
  }
}
```
