---
title: "Mushaf Databases - Comprehensive Statistics Analysis"
author: "Unified Quranic Syntax Adapted Database - UQSAD Project"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

# Mushaf Databases - Comprehensive Statistics Analysis

## Overview

This R Notebook provides comprehensive statistical analysis of all Mushaf databases, focusing on:

-   **Total Ayahs count** across all databases
-   **Words per page statistics** (minimum, maximum, average)
-   **Total pages count** for each Mushaf layout
-   **Lines per page analysis** for different layouts
-   **Surah-to-page mapping** showing which Surahs appear on which pages
-   **Comparative analysis** between different Mushaf layouts and publishers

## Target Databases

The analysis covers **11 different Mushaf databases** with various layouts:

1.  **Indo-Pak 13 lines** (Qudratullah layout)
2.  **QPC v1 15 lines** (Quran Printing Complex version 1)
3.  **Uthmani 15 lines** (Classic Uthmani script)
4.  **Digital Khatt 15 lines** (Modern digital font)
5.  **Indo-Pak 13 lines** (Taj Company)
6.  **QPC Hafs 15 lines** (QPC Hafs version)
7.  **QPC Nastaleeq 15 lines** (Nastaleeq script)
8.  **QPC v2 15 lines** (Quran Printing Complex version 2)
9.  **Qudratullah Indo-Pak 15 lines** (Extended layout)
10. **Taj Indo-Pak 16 lines** (16-line layout)
11. **QPC v4 Tajweed 15 lines** (With Tajweed markings)

## Key Metrics Analyzed

### 📊 Quantitative Metrics

-   Total number of Ayahs in each database
-   Total word count across all Ayahs
-   Complete page counts for each Mushaf
-   Statistical distribution of words per page

### 📏 Layout Metrics

-   Average lines per page for different layouts
-   Comparison between 13, 15, and 16-line layouts
-   Layout efficiency analysis

### 📖 Content Organization

-   Surah distribution across pages
-   Page ranges for each Surah
-   Cross-database consistency analysis

------------------------------------------------------------------------

## Setup and Dependencies

```{r setup, message=FALSE, warning=FALSE}
# Core database connectivity
library(DBI)          # Database interface for R
library(RSQLite)      # SQLite driver for database connections

# Data manipulation and analysis
library(dplyr)        # Data manipulation grammar
library(purrr)        # Functional programming tools

# String processing
library(stringr)      # String manipulation functions

# Visualization and reporting
library(knitr)        # Dynamic report generation
library(tibble)       # Modern data frames

# Optional libraries (will load if available)
if (require(ggplot2, quietly = TRUE)) {
  cat("✅ ggplot2 loaded for visualizations\n")
} else {
  cat("⚠️ ggplot2 not available - visualizations will be skipped\n")
}

if (require(DT, quietly = TRUE)) {
  cat("✅ DT loaded for interactive tables\n")
} else {
  cat("⚠️ DT not available - using basic tables\n")
}

# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8
)

# Record start time for execution tracking
start_time <- Sys.time()

cat("✅ All required libraries loaded successfully\n")
cat("📅 Analysis started at:", format(start_time), "\n")
```

## Database Configuration

```{r database_setup}
# Define the base directory where all Mushaf databases are stored
base_path <- "../data/QUL-Hafs-Mushafs-Mapping"

# Create comprehensive database information table
db_info <- tibble(
  id = 1:11,
  name = c(
    "indopak-13-lines-layout-qudratullah",
    "qpc-v1-15-lines", 
    "uthmani-15-lines",
    "digital-khatt-15-lines",
    "indopak-13-lines-taj-company",
    "qpc-hafs-15-lines",
    "qpc-nastaleeq-15-lines", 
    "qpc-v2-15-lines",
    "qudratullah-indopak-15-lines",
    "taj-indopak-16-lines",
    "qpc-v4-tajweed-15-lines"
  ),
  folder = sprintf("%02d-%s.db", id, name),
  db_file = paste0(name, ".db"),
  full_path = file.path(base_path, folder, db_file),
  display_name = sprintf("%02d - %s", id, name),
  # Extract layout information
  lines_per_page = case_when(
    grepl("13.*lines", name) ~ 13L,
    grepl("15.*lines", name) ~ 15L, 
    grepl("16.*lines", name) ~ 16L,
    TRUE ~ NA_integer_
  ),
  script_type = case_when(
    grepl("indopak", name, ignore.case = TRUE) ~ "Indo-Pak",
    grepl("uthmani", name, ignore.case = TRUE) ~ "Uthmani",
    grepl("nastaleeq", name, ignore.case = TRUE) ~ "Nastaleeq",
    grepl("khatt", name, ignore.case = TRUE) ~ "Digital Khatt",
    TRUE ~ "QPC Standard"
  ),
  publisher = case_when(
    grepl("qpc|hafs", name, ignore.case = TRUE) ~ "QPC",
    grepl("taj", name, ignore.case = TRUE) ~ "Taj Company", 
    grepl("qudratullah", name, ignore.case = TRUE) ~ "Qudratullah",
    TRUE ~ "Other"
  )
)

# Display database configuration
cat("📋 Database Configuration Summary:\n")
print(kable(db_info %>% select(id, display_name, lines_per_page, script_type, publisher), 
            caption = "Mushaf Database Configuration"))
```

## Database Connection Functions

```{r connection_functions}
# Safe database connection with error handling
connect_to_db <- function(db_path) {
  if (file.exists(db_path)) {
    tryCatch({
      conn <- dbConnect(RSQLite::SQLite(), db_path)
      cat("✅ Connected to:", basename(db_path), "\n")
      return(conn)
    }, error = function(e) {
      cat("❌ Connection failed for:", basename(db_path), "-", e$message, "\n")
      return(NULL)
    })
  } else {
    cat("❌ Database file not found:", db_path, "\n")
    return(NULL)
  }
}

# Get comprehensive table information
get_table_info <- function(conn, table_name) {
  if (!is.null(conn) && table_name %in% dbListTables(conn)) {
    tryCatch({
      # Get schema
      schema <- dbGetQuery(conn, paste("PRAGMA table_info(", table_name, ")"))
      # Get row count
      row_count <- dbGetQuery(conn, paste("SELECT COUNT(*) as count FROM", table_name))$count
      # Get sample data
      sample_data <- dbGetQuery(conn, paste("SELECT * FROM", table_name, "LIMIT 3"))
      
      return(list(
        schema = schema,
        row_count = row_count,
        sample_data = sample_data,
        columns = schema$name
      ))
    }, error = function(e) {
      cat("Error getting table info for", table_name, ":", e$message, "\n")
      return(NULL)
    })
  } else {
    return(NULL)
  }
}

cat("🔧 Database connection functions defined\n")
```

## Database Schema Analysis and Pattern Detection

```{r schema_analysis}
# Initialize connection storage
db_connections <- list()
db_schemas <- list()
connection_summary <- tibble()

cat("🔌 Establishing database connections and analyzing schemas...\n\n")

# Connect to all databases and deeply analyze structure
for (i in 1:nrow(db_info)) {
  db_name <- db_info$name[i]
  db_path <- db_info$full_path[i] 
  display_name <- db_info$display_name[i]
  
  cat("Database", i, "of", nrow(db_info), ":", display_name, "\n")
  
  if (file.exists(db_path)) {
    conn <- connect_to_db(db_path)
    
    if (!is.null(conn)) {
      db_connections[[db_name]] <- conn
      tables <- dbListTables(conn)
      
      # Deep schema analysis
      schema_info <- list(
        display_name = display_name,
        tables = tables,
        detailed_schemas = list()
      )
      
      cat("   📋 Tables found:", length(tables), "\n")
      cat("   📋 Table names:", paste(tables, collapse = ", "), "\n")
      
      # Analyze each table in detail
      for (table in tables) {
        tryCatch({
          # Get schema
          table_schema <- dbGetQuery(conn, paste("PRAGMA table_info(", table, ")"))
          # Get sample data (first 3 rows)
          sample_data <- dbGetQuery(conn, paste("SELECT * FROM", table, "LIMIT 3"))
          # Get row count
          row_count <- dbGetQuery(conn, paste("SELECT COUNT(*) as count FROM", table))$count
          
          schema_info$detailed_schemas[[table]] <- list(
            schema = table_schema,
            sample_data = sample_data,
            row_count = row_count,
            columns = table_schema$name
          )
          
          cat("     🔍 Table '", table, "':\n")
          cat("       - Columns:", paste(table_schema$name, collapse = ", "), "\n")
          cat("       - Rows:", row_count, "\n")
          
          # Look for word ID patterns
          word_id_cols <- table_schema$name[grepl("word.*id|id.*word", table_schema$name, ignore.case = TRUE)]
          if (length(word_id_cols) > 0) {
            cat("       - Word ID columns found:", paste(word_id_cols, collapse = ", "), "\n")
            
            # Check for start/end word ID patterns
            for (word_col in word_id_cols) {
              if (nrow(sample_data) > 0 && word_col %in% names(sample_data)) {
                word_range <- dbGetQuery(conn, paste("SELECT MIN(", word_col, ") as min_id, MAX(", word_col, ") as max_id FROM", table))
                cat("       - ", word_col, " range:", word_range$min_id, "to", word_range$max_id, "\n")
              }
            }
          }
          
          # Look for page-related columns
          page_cols <- table_schema$name[grepl("page", table_schema$name, ignore.case = TRUE)]
          if (length(page_cols) > 0) {
            cat("       - Page columns:", paste(page_cols, collapse = ", "), "\n")
          }
          
        }, error = function(e) {
          cat("     ❌ Error analyzing table", table, ":", e$message, "\n")
        })
      }
      
      db_schemas[[db_name]] <- schema_info
      
      connection_summary <- bind_rows(connection_summary, tibble(
        database = display_name,
        status = "✅ Connected & Analyzed",
        table_count = length(tables),
        tables = paste(tables, collapse = ", ")
      ))
      
    } else {
      connection_summary <- bind_rows(connection_summary, tibble(
        database = display_name, 
        status = "❌ Connection Failed",
        table_count = 0,
        tables = "N/A"
      ))
    }
  } else {
    connection_summary <- bind_rows(connection_summary, tibble(
      database = display_name,
      status = "❌ File Not Found", 
      table_count = 0,
      tables = "N/A"
    ))
  }
  cat("\n")
}

cat("📊 Connection and Analysis Summary:\n")
print(kable(connection_summary, caption = "Database Connection and Schema Analysis Status"))
```

## Line-Based Word Analysis (Pages Table)

```{r line_based_analysis}
# Initialize comprehensive statistics storage
mushaf_statistics <- tibble()
page_word_statistics <- tibble()
surah_page_mapping <- tibble()

cat("📈 Starting line-based word analysis from 'pages' table...\n\n")

# Ensure we have database connections from previous step
if (length(db_connections) == 0) {
  cat("❌ No database connections available. Please run the schema analysis section first.\n")
} else {
  cat("✅ Found", length(db_connections), "database connections\n\n")
}

for (db_name in names(db_connections)) {
  conn <- db_connections[[db_name]]
  display_name <- db_info$display_name[db_info$name == db_name]
  
  if (!is.null(conn)) {
    cat("Analyzing:", display_name, "\n")
    
    # Initialize database statistics
    db_stats <- list(
      database = display_name,
      total_words = 0,
      total_lines = 0,
      total_pages = 0,
      total_surahs = 0,
      avg_words_per_page = 0,
      avg_lines_per_page = 0,
      min_words_per_page = Inf,
      max_words_per_page = 0,
      word_id_range = ""
    )
    
    # Focus only on the 'pages' table (skip 'info' table)
    tables <- dbListTables(conn)
    pages_table <- tables[grepl("pages", tables, ignore.case = TRUE)]
    
    if (length(pages_table) > 0) {
      table_name <- pages_table[1]  # Use first pages table found
      
      tryCatch({
        # Get table schema
        schema <- dbGetQuery(conn, paste("PRAGMA table_info(", table_name, ")"))
        columns <- schema$name
        row_count <- dbGetQuery(conn, paste("SELECT COUNT(*) as count FROM", table_name))$count
        
        cat("  📋 Table:", table_name, "with", row_count, "lines\n")
        cat("  📋 Columns:", paste(columns, collapse = ", "), "\n")
        
        # Initialize variables for comparison
        max_word_estimate <- NULL
        
        # METHOD 1: Quick estimate using last_word_id from final row
        if ("last_word_id" %in% columns) {
          max_word_query <- paste("SELECT MAX(CAST(last_word_id AS INTEGER)) as max_word_id FROM", table_name, "WHERE last_word_id IS NOT NULL AND last_word_id != ''")
          max_word_result <- dbGetQuery(conn, max_word_query)
          
          if (nrow(max_word_result) > 0 && !is.na(max_word_result$max_word_id) && !is.null(max_word_result$max_word_id)) {
            max_word_estimate <- as.numeric(max_word_result$max_word_id)
            db_stats$total_words <- max_word_estimate
            cat("  ✓ Quick estimate - Total words (from max last_word_id):", format(db_stats$total_words, big.mark = ","), "\n")
          } else {
            cat("  ⚠️ Quick estimate - No valid last_word_id values found\n")
          }
        }
        
        # METHOD 2: Precise count by summing words per line (using DISTINCT to avoid duplicates)
        if ("first_word_id" %in% columns && "last_word_id" %in% columns) {
          # Calculate words per line: (last_word_id - first_word_id + 1)
          # Use DISTINCT to handle databases with duplicate word ranges
          precise_count_query <- paste("SELECT SUM(CAST(last_word_id AS INTEGER) - CAST(first_word_id AS INTEGER) + 1) as total_words FROM",
                                      "(SELECT DISTINCT first_word_id, last_word_id FROM", table_name, 
                                      "WHERE first_word_id IS NOT NULL AND first_word_id != '' AND last_word_id IS NOT NULL AND last_word_id != '')")
          precise_result <- dbGetQuery(conn, precise_count_query)
          
          if (nrow(precise_result) > 0 && !is.na(precise_result$total_words) && !is.null(precise_result$total_words)) {
            precise_total <- as.numeric(precise_result$total_words)
            cat("  ✓ Precise count - Total words (sum of line calculations):", format(precise_total, big.mark = ","), "\n")
            
            # Use precise count as primary method
            db_stats$total_words <- precise_total
            
            # Compare with quick estimate if available
            if (!is.null(max_word_estimate)) {
              tryCatch({
                if (precise_total != max_word_estimate) {
                  cat("    → Difference between methods:", abs(precise_total - max_word_estimate), "words\n")
                } else {
                  cat("    → Both methods agree!\n")
                }
              }, error = function(e) {
                cat("    → Could not compare methods\n")
              })
            }
          } else {
            cat("  ⚠️ Precise count - No valid word ID values found\n")
          }
        }
        
        # Get word ID range
        if ("first_word_id" %in% columns && "last_word_id" %in% columns) {
          range_query <- paste("SELECT MIN(CAST(first_word_id AS INTEGER)) as min_id, MAX(CAST(last_word_id AS INTEGER)) as max_id FROM", table_name,
                              "WHERE first_word_id IS NOT NULL AND first_word_id != '' AND last_word_id IS NOT NULL AND last_word_id != ''")
          range_result <- dbGetQuery(conn, range_query)
          
          if (nrow(range_result) > 0 && !is.na(range_result$min_id)) {
            db_stats$word_id_range <- paste(range_result$min_id, "to", range_result$max_id)
          }
        }
        
        # Get total lines and pages (using correct column names)
        db_stats$total_lines <- row_count
        
        # Check for page_number column (not just 'page')
        page_col <- NULL
        if ("page_number" %in% columns) {
          page_col <- "page_number"
        } else if ("page" %in% columns) {
          page_col <- "page"
        }
        
        if (!is.null(page_col)) {
          page_count_query <- paste("SELECT MIN(", page_col, ") as min_page, MAX(", page_col, ") as max_page FROM", table_name,
                                   "WHERE", page_col, "IS NOT NULL")
          page_result <- dbGetQuery(conn, page_count_query)
          
          if (nrow(page_result) > 0 && !is.na(page_result$max_page)) {
            db_stats$total_pages <- as.numeric(page_result$max_page)
            cat("  ✓ Pages:", page_result$min_page, "to", page_result$max_page, "\n")
          }
        }
        
        # Calculate words per page (using correct page column and DISTINCT to avoid duplicates)
        if (!is.null(page_col) && "first_word_id" %in% columns && "last_word_id" %in% columns) {
          words_per_page_query <- paste("SELECT", page_col, "as page_num,", 
                                       "SUM(CAST(last_word_id AS INTEGER) - CAST(first_word_id AS INTEGER) + 1) as words",
                                       "FROM (SELECT DISTINCT", page_col, ", first_word_id, last_word_id FROM", table_name,
                                       "WHERE first_word_id IS NOT NULL AND first_word_id != '' AND last_word_id IS NOT NULL AND last_word_id != '')",
                                       "GROUP BY", page_col, "ORDER BY", page_col)
          
          page_words <- dbGetQuery(conn, words_per_page_query)
          
          if (nrow(page_words) > 0) {
            page_words <- page_words %>%
              mutate(database = display_name)
            
            page_word_statistics <- bind_rows(page_word_statistics, page_words)
            
            # Update min/max words per page
            db_stats$min_words_per_page <- min(page_words$words, na.rm = TRUE)
            db_stats$max_words_per_page <- max(page_words$words, na.rm = TRUE)
            
            cat("  ✓ Words per page - Min:", db_stats$min_words_per_page, "Max:", db_stats$max_words_per_page, "\n")
          }
        }
        
        # Calculate lines per page (using correct page column)
        if (!is.null(page_col)) {
          lines_per_page_query <- paste("SELECT AVG(lines_per_page) as avg_lines FROM",
                                       "(SELECT", page_col, ", COUNT(*) as lines_per_page FROM", table_name, 
                                       "WHERE", page_col, "IS NOT NULL GROUP BY", page_col, ")")
          lines_result <- dbGetQuery(conn, lines_per_page_query)
          
          if (nrow(lines_result) > 0 && !is.na(lines_result$avg_lines)) {
            db_stats$avg_lines_per_page <- as.numeric(lines_result$avg_lines)
            cat("  ✓ Average lines per page:", round(db_stats$avg_lines_per_page, 2), "\n")
          }
        }
        
        # Surah analysis - Use surah_name line_type for accurate count
        # Some databases only mark surah numbers on title lines, not ayah lines
        if ("line_type" %in% columns) {
          # Method 1: Count surah_name lines (most reliable)
          surah_count_query <- paste("SELECT COUNT(*) as surah_count FROM", table_name,
                                    "WHERE line_type = 'surah_name'")
          surah_count_result <- dbGetQuery(conn, surah_count_query)
          
          # Handle duplicate surah_name lines (like in QPC Hafs)
          if (nrow(surah_count_result) > 0 && surah_count_result$surah_count > 114) {
            # If more than 114, there might be duplicates - count distinct pages with surah_name
            if (!is.null(page_col)) {
              distinct_surah_pages_query <- paste("SELECT COUNT(DISTINCT", page_col, ") as surah_count FROM", table_name,
                                                "WHERE line_type = 'surah_name' AND", page_col, "IS NOT NULL")
              distinct_result <- dbGetQuery(conn, distinct_surah_pages_query)
              if (nrow(distinct_result) > 0) {
                surah_count_result$surah_count <- distinct_result$surah_count
              }
            }
          }
        } else {
          # Fallback: Use surah_number column if line_type doesn't exist
          surah_cols <- columns[grepl("surah", columns, ignore.case = TRUE)]
          if (length(surah_cols) > 0) {
            surah_col <- surah_cols[1]
            surah_count_query <- paste("SELECT COUNT(DISTINCT", surah_col, ") as surah_count FROM", table_name,
                                      "WHERE", surah_col, "IS NOT NULL AND", surah_col, "> 0")
            surah_count_result <- dbGetQuery(conn, surah_count_query)
          } else {
            surah_count_result <- data.frame(surah_count = 0)
          }
        }
          
          if (nrow(surah_count_result) > 0) {
            db_stats$total_surahs <- as.numeric(surah_count_result$surah_count)
            
            # Surah-to-page mapping - Use surah_name lines for accurate mapping
            if (!is.null(page_col) && "line_type" %in% columns) {
              # Get page numbers where surah_name lines appear (most reliable method)
              mapping_query <- paste("SELECT", 
                                    "ROW_NUMBER() OVER (ORDER BY", page_col, ") as surah_number,",
                                    page_col, "as start_page,", page_col, "as end_page",
                                    "FROM", table_name,
                                    "WHERE line_type = 'surah_name' AND", page_col, "IS NOT NULL",
                                    "ORDER BY", page_col)
              
              surah_mapping <- dbGetQuery(conn, mapping_query)
              if (nrow(surah_mapping) > 0) {
                surah_mapping <- surah_mapping %>% mutate(database = display_name)
                surah_page_mapping <- bind_rows(surah_page_mapping, surah_mapping)
              }
            } else if (!is.null(page_col)) {
              # Fallback to surah_number column if line_type doesn't exist
              surah_cols <- columns[grepl("surah", columns, ignore.case = TRUE)]
              if (length(surah_cols) > 0) {
                surah_col <- surah_cols[1]
                mapping_query <- paste("SELECT", surah_col, "as surah_number, MIN(", page_col, ") as start_page, MAX(", page_col, ") as end_page",
                                      "FROM", table_name, 
                                      "WHERE", surah_col, "IS NOT NULL AND", surah_col, "> 0 AND", page_col, "IS NOT NULL",
                                      "GROUP BY", surah_col, "ORDER BY", surah_col)
                
                surah_mapping <- dbGetQuery(conn, mapping_query)
                if (nrow(surah_mapping) > 0) {
                  surah_mapping <- surah_mapping %>% mutate(database = display_name)
                  surah_page_mapping <- bind_rows(surah_page_mapping, surah_mapping)
                }
              }
            }
            
            cat("  ✓ Surahs:", db_stats$total_surahs, "\n")
          }
        }
        
      }, error = function(e) {
        cat("  ❌ Error analyzing pages table:", e$message, "\n")
      })
    } else {
      cat("  ❌ No 'pages' table found\n")
    }
    
    # Calculate final statistics with error protection
    tryCatch({
      if (!is.null(db_stats$total_pages) && !is.null(db_stats$total_words) && 
          !is.na(db_stats$total_pages) && !is.na(db_stats$total_words) &&
          db_stats$total_pages > 0 && db_stats$total_words > 0) {
        db_stats$avg_words_per_page <- round(db_stats$total_words / db_stats$total_pages, 2)
      } else {
        db_stats$avg_words_per_page <- 0
      }
      
      # Handle infinite values
      if (is.infinite(db_stats$min_words_per_page) || is.na(db_stats$min_words_per_page)) {
        db_stats$min_words_per_page <- 0
      }
      if (is.infinite(db_stats$max_words_per_page) || is.na(db_stats$max_words_per_page)) {
        db_stats$max_words_per_page <- 0
      }
      
      # Ensure all numeric fields are properly set
      numeric_fields <- c("total_words", "total_lines", "total_pages", "total_surahs", 
                         "avg_words_per_page", "avg_lines_per_page", "min_words_per_page", "max_words_per_page")
      for (field in numeric_fields) {
        if (is.null(db_stats[[field]]) || is.na(db_stats[[field]])) {
          db_stats[[field]] <- 0
        }
      }
      
      # Ensure character fields are set
      if (is.null(db_stats$word_id_range) || is.na(db_stats$word_id_range)) {
        db_stats$word_id_range <- ""
      }
      
      # Add to comprehensive statistics
      mushaf_statistics <- bind_rows(mushaf_statistics, as_tibble(db_stats))
      
      cat("  📊 Final Summary:")
      cat(" Words:", format(db_stats$total_words, big.mark = ","))
      cat(" | Pages:", db_stats$total_pages)  
      cat(" | Lines:", db_stats$total_lines)
      cat(" | Range:", db_stats$word_id_range, "\n\n")
      
    }, error = function(e) {
      cat("  ❌ Error in final calculations:", e$message, "\n\n")
    })
  }
}

cat("✅ Line-based word analysis completed for all databases\n")
```

## Summary Statistics Dashboard

```{r statistics_dashboard}
cat("📊 MUSHAF DATABASES - COMPREHENSIVE STATISTICS DASHBOARD\n")
cat("="=rep("=", 60), "\n\n")

# Overall Statistics Table
cat("### Overall Statistics Summary\n")
if (nrow(mushaf_statistics) > 0) {
  summary_table <- mushaf_statistics %>%
    select(database, total_words, total_pages, total_lines, total_surahs, avg_words_per_page, avg_lines_per_page, word_id_range) %>%
    arrange(desc(total_words))

  print(kable(summary_table, 
              caption = "Complete Statistics for All Mushaf Databases",
              col.names = c("Database", "Total Words", "Total Pages", "Total Lines", "Total Surahs", "Avg Words/Page", "Avg Lines/Page", "Word ID Range")))
} else {
  cat("❌ No statistics data available. Please check the line-based analysis section.\n")
}

# Words per Page Distribution
if (nrow(page_word_statistics) > 0) {
  cat("\n### Words per Page Analysis\n")
  page_summary <- page_word_statistics %>%
    group_by(database) %>%
    summarise(
      total_pages = n(),
      min_words = min(words, na.rm = TRUE),
      max_words = max(words, na.rm = TRUE), 
      avg_words = round(mean(words, na.rm = TRUE), 2),
      median_words = round(median(words, na.rm = TRUE), 2),
      std_dev = round(sd(words, na.rm = TRUE), 2),
      .groups = 'drop'
    ) %>%
    arrange(desc(avg_words))
  
  print(kable(page_summary,
              caption = "Words per Page Distribution Statistics",
              col.names = c("Database", "Pages", "Min Words", "Max Words", "Avg Words", "Median Words", "Std Dev")))
}

# Layout Comparison
cat("\n### Layout Comparison Analysis\n")
layout_comparison <- mushaf_statistics %>%
  left_join(db_info %>% select(display_name, lines_per_page, script_type, publisher), 
            by = c("database" = "display_name")) %>%
  select(database, lines_per_page, script_type, publisher, total_pages, avg_words_per_page, avg_lines_per_page) %>%
  arrange(lines_per_page, script_type)

print(kable(layout_comparison,
            caption = "Layout and Script Type Comparison",
            col.names = c("Database", "Lines/Page", "Script Type", "Publisher", "Total Pages", "Avg Words/Page", "Avg Lines/Page")))
```

## Data Visualizations

```{r visualizations, fig.width=12, fig.height=10}
# Create comprehensive visualization dashboard
if (require(ggplot2, quietly = TRUE) && require(gridExtra, quietly = TRUE) && nrow(mushaf_statistics) > 0) {
  
  # 1. Total Words Comparison
  p1 <- ggplot(mushaf_statistics, aes(x = reorder(database, total_words), y = total_words)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    coord_flip() +
    labs(title = "Total Words Across Mushaf Databases", 
         x = NULL, y = "Total Words") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
  
  # 2. Total Pages Comparison  
  p2 <- ggplot(mushaf_statistics, aes(x = reorder(database, total_pages), y = total_pages)) +
    geom_col(fill = "darkgreen", alpha = 0.8) +
    coord_flip() +
    labs(title = "Total Pages Across Mushaf Databases",
         x = NULL, y = "Total Pages") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
  
  # 3. Average Words per Page
  stats_with_words <- mushaf_statistics %>% filter(avg_words_per_page > 0)
  if (nrow(stats_with_words) > 0) {
    p3 <- ggplot(stats_with_words, aes(x = reorder(database, avg_words_per_page), y = avg_words_per_page)) +
      geom_col(fill = "orange", alpha = 0.8) +
      coord_flip() +
      labs(title = "Average Words per Page",
           x = NULL, y = "Average Words per Page") +
      theme_minimal() +
      theme(axis.text.y = element_text(size = 8))
  } else {
    p3 <- NULL
  }
  
  # 4. Words per Page Distribution (if data available)
  if (nrow(page_word_statistics) > 0 && nrow(page_word_statistics) < 5000) {  # Avoid plotting too much data
    p4 <- ggplot(page_word_statistics, aes(x = words, fill = database)) +
      geom_histogram(alpha = 0.7, bins = 30, position = "identity") +
      facet_wrap(~database, scales = "free_y", ncol = 2) +
      labs(title = "Distribution of Words per Page by Database",
           x = "Words per Page", y = "Frequency") +
      theme_minimal() +
      theme(legend.position = "none",
            strip.text = element_text(size = 8),
            axis.text = element_text(size = 7))
    
    # Combine all plots (including p4)
    if (!is.null(p3)) {
      grid.arrange(p1, p2, p3, p4, ncol = 2, heights = c(1, 1))
    } else {
      grid.arrange(p1, p2, p4, ncol = 2, heights = c(1, 1))
    }
  } else {
    # Just show first plots without p4
    if (!is.null(p3)) {
      grid.arrange(p1, p2, p3, ncol = 2, heights = c(1, 1))
    } else {
      grid.arrange(p1, p2, ncol = 1)
    }
  }
  
} else {
  if (nrow(mushaf_statistics) == 0) {
    cat("📊 No statistics data available for visualization. Please check the analysis sections.\n")
  } else {
    cat("📊 ggplot2 or gridExtra not available. Install with: install.packages(c('ggplot2', 'gridExtra'))\n")
  }
}
```

## Surah-to-Page Mapping Analysis

```{r surah_mapping_analysis}
if (nrow(surah_page_mapping) > 0) {
  cat("📖 SURAH-TO-PAGE MAPPING ANALYSIS\n")
  cat("="=rep("=", 40), "\n\n")
  
  # Clean and standardize surah mapping data
  surah_clean <- surah_page_mapping %>%
    mutate(
      # Clean surah identifiers (using only surah_number since surah_name doesn't exist)
      # Filter out Surah 0 as it doesn't exist (Quran has Surahs 1-114)
      surah_id = case_when(
        !is.na(surah_number) & surah_number > 0 ~ paste("Surah", surah_number),
        TRUE ~ "Unknown"
      ),
      # Extract page ranges (using existing column names)
      page_start = case_when(
        !is.na(start_page) ~ start_page,
        TRUE ~ NA_integer_
      ),
      page_end = case_when(
        !is.na(end_page) ~ end_page,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(page_start) & surah_number > 0) %>%
    arrange(database, page_start)
  
  # Sample of surah mappings
  cat("### Sample Surah-to-Page Mappings\n")
  sample_mapping <- surah_clean %>%
    select(database, surah_id, page_start, page_end) %>%
    group_by(database) %>%
    slice_head(n = 10) %>%
    ungroup()
  
  if (nrow(sample_mapping) > 0) {
    print(kable(sample_mapping,
                caption = "Sample Surah-to-Page Mappings (First 10 per database)",
                col.names = c("Database", "Surah", "Start Page", "End Page")))
  }
  
  # Summary statistics for surah mappings
  cat("\n### Surah Mapping Summary Statistics\n")
  mapping_summary <- surah_clean %>%
    group_by(database) %>%
    summarise(
      total_surahs = n(),
      avg_pages_per_surah = round(mean(page_end - page_start + 1, na.rm = TRUE), 2),
      min_pages_per_surah = min(page_end - page_start + 1, na.rm = TRUE),
      max_pages_per_surah = max(page_end - page_start + 1, na.rm = TRUE),
      .groups = 'drop'
    )
  
  print(kable(mapping_summary,
              caption = "Surah Distribution Statistics by Database",
              col.names = c("Database", "Total Surahs", "Avg Pages/Surah", "Min Pages/Surah", "Max Pages/Surah")))
  
} else {
  cat("📖 No surah-to-page mapping data available in the analyzed databases.\n")
}
```

## Detailed Database Reports

```{r detailed_reports}
cat("📋 DETAILED DATABASE REPORTS\n")
cat("="=rep("=", 50), "\n\n")

for (i in 1:nrow(mushaf_statistics)) {
  stats <- mushaf_statistics[i, ]
  db_details <- db_info %>% filter(display_name == stats$database)
  
  cat("## Database:", stats$database, "\n")
  cat(rep("-", nchar(stats$database) + 15), "\n")
  
  cat("### 📊 Quantitative Metrics\n")
  cat("• Total Words:", format(stats$total_words, big.mark = ","), "\n") 
  cat("• Total Lines:", format(stats$total_lines, big.mark = ","), "\n")
  cat("• Total Pages:", format(stats$total_pages, big.mark = ","), "\n")
  cat("• Total Surahs:", stats$total_surahs)
  
  # Add note for databases with incomplete surah data
  if (grepl("nastaleeq", stats$database, ignore.case = TRUE) && stats$total_surahs == 1) {
    cat(" (Note: Surah data incomplete - database contains full Quran text but lacks proper surah markers)")
  } else if (stats$total_surahs == 1) {
    cat(" (Note: Database appears to have incomplete surah data)")
  }
  cat("\n")
  
  cat("• Word ID Range:", stats$word_id_range, "\n")
  
  cat("\n### 📏 Page Metrics\n")
  cat("• Average Words per Page:", stats$avg_words_per_page, "\n")
  cat("• Average Lines per Page:", round(stats$avg_lines_per_page, 1), "\n")
  if (stats$min_words_per_page > 0) {
    cat("• Minimum Words per Page:", stats$min_words_per_page, "\n")
    cat("• Maximum Words per Page:", stats$max_words_per_page, "\n")
  }
  
  if (nrow(db_details) > 0) {
    cat("\n### 📖 Layout Information\n")
    if (!is.na(db_details$lines_per_page)) {
      cat("• Lines per Page:", db_details$lines_per_page, "\n")
    }
    cat("• Script Type:", db_details$script_type, "\n")
    cat("• Publisher:", db_details$publisher, "\n")
  }
  
  # Add page statistics if available
  db_page_stats <- page_word_statistics %>% filter(database == stats$database)
  if (nrow(db_page_stats) > 0) {
    cat("\n### 📊 Words per Page Distribution\n")
    cat("• Standard Deviation:", round(sd(db_page_stats$words, na.rm = TRUE), 2), "\n")
    cat("• Median Words per Page:", median(db_page_stats$words, na.rm = TRUE), "\n")
    cat("• Pages Analyzed:", nrow(db_page_stats), "\n")
  }
  
  cat("\n")
}
```

## Data Export and Summary

```{r data_export}
# Create comprehensive summary for export
final_summary <- mushaf_statistics %>%
  left_join(db_info %>% select(display_name, lines_per_page, script_type, publisher), 
            by = c("database" = "display_name")) %>%
  arrange(desc(total_ayahs))

# Export option (uncomment to save to CSV)
# write.csv(final_summary, "mushaf_statistics_summary.csv", row.names = FALSE)
# write.csv(page_statistics, "words_per_page_statistics.csv", row.names = FALSE)
# write.csv(surah_page_mapping, "surah_page_mappings.csv", row.names = FALSE)

cat("📊 FINAL SUMMARY\n")
cat("="=rep("=", 30), "\n")
cat("✅ Databases analyzed:", nrow(final_summary), "\n")
cat("📖 Total unique Mushaf layouts:", length(unique(final_summary$lines_per_page)), "\n")
cat("🖋️ Script types covered:", length(unique(final_summary$script_type)), "\n")
cat("🏢 Publishers analyzed:", length(unique(final_summary$publisher)), "\n")

if (nrow(page_statistics) > 0) {
  cat("📄 Total pages analyzed:", nrow(page_statistics), "\n")
}

if (nrow(surah_page_mapping) > 0) {
  cat("📚 Surah mappings extracted:", nrow(surah_page_mapping), "\n")
}

cat("\n🎯 Analysis completed successfully!\n")
```

## Database Connection Cleanup

```{r cleanup}
# Properly close all database connections
for (conn in db_connections) {
  if (!is.null(conn)) {
    dbDisconnect(conn)
  }
}

cat("🔐 All database connections closed successfully.\n")
```

------------------------------------------------------------------------

**Analysis completed:** `r Sys.time()`\
**Total execution time:** `r Sys.time() - start_time` (if start_time was defined)

------------------------------------------------------------------------
